Increasing discovery rates in preclinical research through optimised statistical decision criteria

### 2021-02-03 Manuscript/MS_Nature_Methods.Rmd
changed figure Nat Met MS

---
title: 'MS Nature Methods'
csl: nature-methods.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 11pt
header-includes:
       - \usepackage{subfig}
abstract: \singlespacing Brief Communications begin with a brief unreferenced abstract (3 sentences, no more than 70 words),
bibliography: refs_repro.bib
---

@@

\ \ \ \ Preclinical research is essential for identifying promising interventions and to generate robust evidence to support translation to humans. Experiments in a preclinical setting are conducted in two different operating modes [@kimmelman2014distinguishing]. Early-stage preclinical experiments are *exploratory* with the aim to discover potentially effective interventions and generate hypotheses. These are tested at a later stage under more strict conditions in *confirmatory* mode [@landis2012call].

Various publications have called for a distinction between exploration and confirmation in preclinical animal research [@dirnagl2016thomas; @kimmelman2014distinguishing; @landis2012call]. Mogil & Macleod [@mogil2017no] took up this sentiment and proposed that an initial claim must be independently confirmed in order to be published. This suggestion emphasizes confirmation of initial findings. However, the question remains which exploratory results should be contested during confirmation. There is currently no guidance regarding criteria that should be applied to make this decision.

Commonly, the *p*-value is used to make a decision whether or not to further investigate a claim. Given that prior probabilities at this stage of discovery are low and sample sizes are small [@macleod2008evidence; @howells2014bringing], the *p*-value might prematurely discard potentially promising interventions [@dirnagl2020resolving].

Exploration requires sensitive tests to detect rare and possibly small effects [@bonapersona2019repair; @dirnagl2020resolving]. As more sensitive criteria invite more false positive results, confirmation must aim at reducing false positives to ensure that only true effects are carried forward to clinical testing. To increase power and safeguard reliability of results, sample sizes must be increased when switching from exploratory to confirmatory mode. However, ethical, time, and budget constraints prevent preclinical researchers from increasing their sample size and thus complicate the detection and precise estimation of effects. In an effort to prevent false negative results in exploration and reduce false positives during confirmation, it is necessary to devise strategies to move from exploration to confirmation that meet these complementary goals.

Here, we consider how to move from exploratory to confirmatory mode while balancing the number of animals against the likelihood of false negative and false positive outcomes.

To this end, we simulated two preclinical research trajectories comprising an exploratory study and a first confirmatory study (Figure 1?).

@@

We based our simulations on two empirical effect size distributions reflecting an *optimistic* (high pre-study odds) and a *pessimistic* (low pre-study odds) scenario (Supplementary Figure 1). After an initial exploratory study, a first decision identified experiments that should move from exploratory to confirmatory mode. One trajectory (standard) employed the conventional significance threshold ($\alpha$ = .05) for this decision. The second trajectory (SESOI) used a more lenient threshold based on an *a priori* determined smallest effect size of interest (SESOI).

In the standard trajectory, in the optimistic scenario, `r exploratory_data_sig_Szucs_10_0.05$N / 100` percent of experiments met the criterion *p* $\leq$ .05. In the pessimistic scenario, `r exploratory_data_sig_Carneiro_10_0.05$N / 100` percent of experiments had a *p*-value $\leq$ .05. A closer look at the effect sizes of the experiments that proceeded to confirmation reveals that the conventional significance threshold is a conservative filter. Many effect sizes $\neq$ 0 are eliminated and not further investigated (Figure \@ref(fig:histograms)a--b). This demonstrates that the conventional significance threshold is not useful to screen for potentially meaningful effects at the early stages of drug discovery even if pre-study odds are high.

In the SESOI trajectory, we estimated the exploratory effect size and 95 percent confidence interval (CI) around that estimate. We examined whether the CI covered our SESOI (Cohen's *d* of 0.5 and 1.0, respectively). If this was the case for an experiment, it advanced to confirmatory mode. Importantly, we did not consider the *p*-value additionally. Applying this decision criterion resulted in `r exploratory_data_Szucs_10_0.5$N / 100` and `r exploratory_data_Szucs_10_1.0$N / 100` percent of experiments moving to confirmation in case of the optimistic distribution and `r exploratory_data_Carneiro_10_0.5$N / 100` and `r exploratory_data_Carneiro_10_1.0$N / 100` percent for the pessimistic distribution based on a SESOI of 0.5 and 1.0, respectively. Compared to the conventional significance threshold, the range of effect size estimates that proceeded to confirmation shifted and included less extreme values, as well as more values closer to zero (Figure \@ref(fig:histograms)c--f). This decision criterion meets the goal of preventing false negatives better than the conventional significance threshold.

@@

@@

Overall, our simulation shows that current practice reflected by the standard trajectory does not meet the complementary goals of exploration and confirmation along the preclinical research trajectory. In the standard trajectory, the switch from exploratory to confirmatory mode is extremely costly, as it eliminates numerous potentially meaningful effects. This shortcoming is also reflected in the PPV where even in the best-case scenario only 40 out 100 significant results reflect an underlying true effect (compared to 73/100 in the SESOI trajectory). In light of our findings, we advise a rejection of the standard approach. We present here an easily applicable alternative to current practice. Optimizing decision criteria and sample size calculations by employing a SESOI increases chances to detect true effects.

@@

>>>>

---
title: 'Increasing discovery rates in preclinical research through optimised statistical decision criteria'
csl: nature-methods.csl
indent: yes
output: bookdown::pdf_document2
toc: false
linestretch: 1.5
fontsize: 11pt
header-includes:
       - \usepackage{subfig}
         \usepackage{ulem}
abstract: \singlespacing Brief Communications begin with a brief unreferenced abstract (3 sentences, no more than 70 words),
bibliography: refs_repro.bib
---

@@

\ \ \ \ Preclinical research is essential for identifying promising interventions and to generate robust evidence to support translation to humans. To achieve these goals, experiments are conducted in two different operating modes [@kimmelman2014distinguishing]. Early-stage preclinical experiments are *exploratory* with the aim to discover potentially effective interventions and generate hypotheses. These are tested at a later stage under more strict conditions in *confirmatory* mode [@landis2012call].

This sequential approach will increase the likelihood that generated evidence is robust and potentially reduce translational failure [@dirnagl2016thomas; @kimmelman2014distinguishing; @landis2012call;@mogil2017no][ADD elife paper here and cut landis and kimmelman]. Before conducting confirmatory experiments, researchers need to decide which exploratory results should be contested. Commonly, such a decision is based on a statistical test and associated *p*-value threshold in an exploratory study [do we have a good citation here?]. Here, we argue that this standard approach is detrimental to the goal to actually identify efficacious treatments. We propose an alternative that is, mimicking clinical trials, focused on a smallest effect size of interest. We show through simulations that this will increase the probability that a confirmatory study identifies a true effect, i.e. the positive predictive value.[perhaps cite Ionanidis 2005 here]. 
[and will increaeGiven that prior probabilities at this stage of discovery are low and sample sizes are small [@macleod2008evidence; @howells2014bringing], the *p*-value might prematurely discard potentially promising interventions [@dirnagl2020resolving].]

Our approach is based on the notion that exploration and confirmation fulfill different roles in scientific discovery. Exploration detects via sensitive tests potentially relevant effects among many hypotheses that have a low prior probability of being true (in the following \textit{pre-study odds}). [@bonapersona2019repair; @dirnagl2020resolving]. As more sensitive criteria invite more false positive results, confirmation must aim at reducing false positives to ensure that only true effects are carried forward to subsequent testing. [To increase power and safeguard reliability of results, sample sizes must be increased when switching from exploratory to confirmatory mode. suggest delete] To complicate matters, ethical, time, and budget constraints limit degrees of freedom in experimental design. Consequently, to prevent false negative results in exploration and to reduce false positives during confirmation, it is necessary to devise strategies to move from exploration to confirmation that meet these complementary goals.

Specifically, the likelihood of false negative and false positive outcomes needs to be balanced against the number of animals. [Perhaps can give a small example here if space allows]

To this end, we simulated two preclinical research trajectories each comprising an exploratory study and a first confirmatory study (Figure 1?).

@@

We based our simulations on two published empirical effect size distributions.  reflecting an *optimistic* (high pre-study odds) and a *pessimistic* (low pre-study odds) scenario (Supplementary Figure 1). After an initial exploratory study, a decision criterion identified experiments that should move from exploratory to confirmatory mode. One trajectory (standard) employed the conventional significance threshold ($\alpha$ = .05) for this decision. The second trajectory (SESOI) used a more lenient threshold based on an *a priori* determined smallest effect size of interest (SESOI), a practice also employed in clinical trials [CITATION]. [copied from below] For this, we estimated the effect size of each exploratory study and associated 95 percent confidence interval (CI). We examined whether this CI covered our SESOI (Cohen's *d* of 0.5 and 1.0, respectively).

\sout{In the standard trajectory, under optimistic pre-study odds, `r exploratory_data_sig_Szucs_10_0.05$N / 100` percent of experiments met the criterion *p* $\leq$ .05. In the pessimistic scenario, `r exploratory_data_sig_Carneiro_10_0.05$N / 100` percent of experiments had a *p*-value $\leq$ .05. Effect sizes of experiments that proceeded to confirmation reveals that the conventional significance threshold is a conservative filter. Many effect sizes $\neq$ 0 are eliminated and not further investigated (Figure \@ref(fig:histograms)a--b). This demonstrates that the conventional significance threshold is not useful to screen for potentially meaningful effects at the early stages of drug discovery even if pre-study odds are high.}

\emph{I rephrased the entire previous paragraph for readability. The main take home message in this rewrite is that you currently do not guide the reader through the results. You state the results but do not imply an interpretation.}

As expected, more experiments transitioned to the confirmatory phase in the SESOI trajectory compared to the standard trajectory. This was true for both the optimistic and pessimistic pre-study odds (PUT STATS HERE). This was also reflected in effect sizes where many effect sizes $\neq$ 0 are eliminated and not further investigated (Figure \@ref(fig:histograms)a--b). This demonstrates that the conventional significance threshold is not useful to screen for potentially meaningful effects at the early stages of drug discovery even if pre-study odds are high. Contrary, the range of effect size estimates in the SESOI trajectory that proceeded to confirmation shifted[shifted where?]  and included less extreme values, as well as more values closer to zero (Figure \@ref(fig:histograms)c--f). This decision criterion thus reduces false negatives compared to the conventional significance threshold.

\sout{In the SESOI trajectory, we estimated the exploratory effect size and 95 percent confidence interval (CI) around that estimate. We examined whether the CI covered our SESOI (Cohen's *d* of 0.5 and 1.0, respectively). If this was the case for an experiment, it advanced to confirmatory mode. Importantly, we did not consider the *p*-value additionally. Applying this decision criterion resulted in `r exploratory_data_Szucs_10_0.5$N / 100` and `r exploratory_data_Szucs_10_1.0$N / 100` percent of experiments moving to confirmation in case of the optimistic distribution and `r exploratory_data_Carneiro_10_0.5$N / 100` and `r exploratory_data_Carneiro_10_1.0$N / 100` percent for the pessimistic distribution based on a SESOI of 0.5 and 1.0, respectively. Compared to the conventional significance threshold, the range of effect size estimates that proceeded to confirmation shifted and included less extreme values, as well as more values closer to zero (Figure \@ref(fig:histograms)c--f). This decision criterion meets the goal of preventing false negatives better than the conventional significance threshold.}

@@

[Please adjust the following paragraphs in the same way I have adjusted the one above. State the result and give descriptive stats in brackets. As it is this is way too long and does not read smoothly.]

@@

@@

Overall, our simulation shows that current practice reflected by the standard trajectory does not meet the complementary goals of exploration and confirmation along a preclinical research trajectory. In the standard trajectory, the switch from exploratory to confirmatory mode eliminates numerous potentially meaningful effects. This shortcoming is also reflected in the PPV where only 40 out 100 significant results reflect an underlying true effect (compared to 73/100 in the SESOI trajectory). In light of our findings, we advise a rejection of the standard approach. [After shortening the results part significantly here we can drop a few lines more on how to apply this. Adjust the SESOI to the feasibility in the lab. This will raise awareness for the limitations of the study! We could also hint at the number of initial animals and how these influence our results. We should also emphasise the deviation from one experiment one result practice but towards a systemic investigation of true effects. Also hint at the possibility for multi center trials and how much costs are saved downstream].We present here an easily applicable alternative to current practice. Optimizing decision criteria and sample size calculations by employing a SESOI increases chances to detect true effects while keeping the number of animals reasonably low.
